### 2022-02-28 doc/src/DuoPLIC.tex
Adopt "APLIC" as common name for Advanced PLIC

The Advanced PLIC of the previous chapter is not backward compatible
with the original {\RISCV} PLIC.%
\footnote{%
A proposed standard for the original PLIC is the document
\textit{{\RISCV} Platform-Level Interrupt Controller Specification},
available as of mid-2021 here:
\z{https://github.com/{\allowbreak}riscv/{\allowbreak}riscv-plic-spec}.
However, both free implementations of the PLIC and those of some major
{\RISCV} vendors such as SiFive have been observed to differ in small
but significant ways from this document and from each other.
Some important details of the original PLIC were better documented in
older versions of
\textit{The {\RISCV} Instruction Set Manual, Volume~II: Privileged
Architecture}, in a chapter titled
``Platform-Level Interrupt Controller (PLIC).''
That chapter has since been removed from more recent versions of
\textit{The {\RISCV} Instruction Set Manual}.%
}
To ease the transition of software from the original PLIC to the
Advanced PLIC, a {\RISCV} system may have a \emph{\mbox{Duo-PLIC\/}}
that software can configure to act as either form of PLIC.
A \mbox{Duo-PLIC} is an extension of an Advanced PLIC, having all the
usual memory-mapped interfaces of an Advanced PLIC, plus, at some other
memory addresses, the memory-mapped interface of an original {\RISCV}
PLIC too.
When the \mbox{Duo-PLIC} has \emph{compatibility mode} enabled,
interrupts from the \mbox{Duo-PLIC} to harts are controlled exclusively
by the original PLIC side of the \mbox{Duo-PLIC};
and when compatibility mode is disabled, interrupts to harts
are controlled exclusively by the Advanced PLIC side of the
\mbox{Duo-PLIC}.
A \mbox{Duo-PLIC} comes out of reset with compatibility mode enabled,
so older software with no knowledge of an Advanced PLIC can see the
\mbox{Duo-PLIC} as an original PLIC.

When compatibility mode is enabled so a \mbox{Duo-PLIC} acts as an
original PLIC, the \mbox{Duo-PLIC} supports only the direct delivery of
interrupts to harts, not the forwarding of interrupts by MSIs.
When compatibility mode is disabled and the \mbox{Duo-PLIC} acts as
an Advanced PLIC, direct delivery mode is supported for all interrupt
domains, and MSI delivery mode may optionally be supported.
Hence, for each Advanced PLIC interrupt domain, field DM
(Delivery Mode) of the \z{domaincfg} register is either read-only zero
or writable, but is not read-only one.

In the memory-mapped control region for the Advanced PLIC root
interrupt domain (only), a \mbox{Duo-PLIC} adds to the \z{domaincfg}
register a CM (Compatibility Mode) field in bit~6.
The complete format of the root interrupt domain's \z{domaincfg} is
thus:
\begin{displayLinesTable}[l@{\quad}l]
bits 31:24 & read-only \z{0x80} \\
bit 8      & IE \\
bit 7      & read-only 0 \\
bit 6      & CM \\
bit 2      & DM (\WARL) \\
bit 0      & BE (\WARL) \\
\end{displayLinesTable}
All other register bits are reserved and read as zeros.

Reset initializes the Advanced PLIC root domain's \z{domaincfg}
register with CM =~1 and with all other writable bits zeros.

@@

\item
For each \emph{context} implemented by the original PLIC side of the
\mbox{Duo-PLIC}, the output interrupt for that context becomes the
interrupt signal delivered from the \mbox{Duo-PLIC} to the specific
hart and privilege level specified by the implementation.
Other than the CM field of the Advanced PLIC root domain's
\z{domaincfg} register, the state of the Advanced PLIC side of the
\mbox{Duo-PLIC} is irrelevant to the interrupts delivered to harts.

\item
In the memory-mapped control region for each Advanced PLIC interrupt
domain:
\begin{itemize}

@@

\item
Interrupts to harts are determined by the Advanced PLIC side of the
\mbox{Duo-PLIC}, as specified in Chapter~\ref{ch:AdvPLIC}.
The state of the original PLIC side of the \mbox{Duo-PLIC} is
irrelevant.

@@

If the \mbox{Duo-PLIC}'s compatibility mode is changed from
enabled to disabled, the registers of the Advanced PLIC side of
the \mbox{Duo-PLIC} obtain valid and consistent but otherwise
{\unspecified} values, except for the \z{domaincfg} registers
and, if they exist, the registers that configure MSI addresses
(\z{mmsiaddrcfg}, \z{mmsiaddrcfgh}, \z{smsiaddrcfg}, and
\z{smsiaddrcfgh}), all of which retain their previous values.
(The root domain's \z{domaincfg} gets the value written when changing
field~CM, of course.)

The endianness of the memory-mapped registers of the original PLIC side
of the \mbox{Duo-PLIC} is not affected by the BE field of any Advanced
PLIC \z{domaincfg} register.
For most, maybe all versions of an original PLIC, the registers are
always little-endian.

The number of interrupt priority levels supported by the original PLIC
side of a \mbox{Duo-PLIC} is the same as that of the Advanced PLIC side
when in direct delivery mode.
Hence, the number of integer bits to store a priority number, IPRIOLEN,
is a constant for a \mbox{Duo-PLIC} (in the range 1 to 8 as usual),
independent of whether compatibility mode is enabled or disabled.

However, an original PLIC encodes priorities in the opposite order
from an IMSIC or Advanced PLIC, with number 1 being interpreted as the
lowest priority and larger numbers as higher priorities.
When a \mbox{Duo-PLIC} is acting as an original PLIC and reports to
a hart an interrupt priority number (see Chapter~\ref{ch:MSLevel}
concerning a hart's handling of interrupt priorities), the number
reported is negated to \emph{normalize} it to the convention of the
Advanced Interrupt Architecture.
If the original PLIC's priority number for an interrupt is~$p$,
the normalized priority number communicated to the target hart is
${\mbox{2}^\textrm{IPRIOLEN}-p}$.

>>>>

The Advanced PLIC of the previous chapter is not backward compatible
with the original {\RISCV} PLIC.%
\footnote{%
A proposed standard for the original PLIC is the document
\textit{{\RISCV} Platform-Level Interrupt Controller Specification},
available as of mid-2021 here:
\z{https://github.com/{\allowbreak}riscv/{\allowbreak}riscv-plic-spec}.
However, both free implementations of the PLIC and those of some major
{\RISCV} vendors such as SiFive have been observed to differ in small
but significant ways from this document and from each other.
Some important details of the original PLIC were better documented in
older versions of
\textit{The {\RISCV} Instruction Set Manual, Volume~II: Privileged
Architecture}, in a chapter titled
``Platform-Level Interrupt Controller (PLIC).''
That chapter has since been removed from more recent versions of
\textit{The {\RISCV} Instruction Set Manual}.%
}
To ease the transition of software from the original PLIC to the
APLIC, a {\RISCV} system may have a \emph{\mbox{Duo-PLIC\/}}
that software can configure to act as either form of PLIC.
A \mbox{Duo-PLIC} is an extension of an APLIC, having all the
usual memory-mapped interfaces of an APLIC, plus, at some other
memory addresses, the memory-mapped interface of an original {\RISCV}
PLIC too.
When the \mbox{Duo-PLIC} has \emph{compatibility mode} enabled,
interrupts from the \mbox{Duo-PLIC} to harts are controlled exclusively
by the original PLIC side of the \mbox{Duo-PLIC};
and when compatibility mode is disabled, interrupts to harts
are controlled exclusively by the APLIC side of the
\mbox{Duo-PLIC}.
A \mbox{Duo-PLIC} comes out of reset with compatibility mode enabled,
so older software with no knowledge of an APLIC can see the
\mbox{Duo-PLIC} as an original PLIC.

When compatibility mode is enabled so a \mbox{Duo-PLIC} acts as an
original PLIC, the \mbox{Duo-PLIC} supports only the direct delivery of
interrupts to harts, not the forwarding of interrupts by MSIs.
When compatibility mode is disabled and the \mbox{Duo-PLIC} acts as
an APLIC, direct delivery mode is supported for all interrupt
domains, and MSI delivery mode may optionally be supported.
Hence, for each APLIC interrupt domain, field DM
(Delivery Mode) of the \z{domaincfg} register is either read-only zero
or writable, but is not read-only one.

In the memory-mapped control region for the APLIC root
interrupt domain (only), a \mbox{Duo-PLIC} adds to the \z{domaincfg}
register a CM (Compatibility Mode) field in bit~6.
The complete format of the root interrupt domain's \z{domaincfg} is
thus:
\begin{displayLinesTable}[l@{\quad}l]
bits 31:24 & read-only \z{0x80} \\
bit 8      & IE \\
bit 7      & read-only 0 \\
bit 6      & CM \\
bit 2      & DM (\WARL) \\
bit 0      & BE (\WARL) \\
\end{displayLinesTable}
All other register bits are reserved and read as zeros.

Reset initializes the APLIC root domain's \z{domaincfg}
register with CM =~1 and with all other writable bits zeros.

@@

\item
For each \emph{context} implemented by the original PLIC side of the
\mbox{Duo-PLIC}, the output interrupt for that context becomes the
interrupt signal delivered from the \mbox{Duo-PLIC} to the specific
hart and privilege level specified by the implementation.
Other than the CM field of the APLIC root domain's
\z{domaincfg} register, the state of the APLIC side of the
\mbox{Duo-PLIC} is irrelevant to the interrupts delivered to harts.

\item
In the memory-mapped control region for each APLIC interrupt
domain:
\begin{itemize}

@@

\item
Interrupts to harts are determined by the APLIC side of the
\mbox{Duo-PLIC}, as specified in Chapter~\ref{ch:AdvPLIC}.
The state of the original PLIC side of the \mbox{Duo-PLIC} is
irrelevant.

@@

If the \mbox{Duo-PLIC}'s compatibility mode is changed from
enabled to disabled, the registers of the APLIC side of
the \mbox{Duo-PLIC} obtain valid and consistent but otherwise
{\unspecified} values, except for the \z{domaincfg} registers
and, if they exist, the registers that configure MSI addresses
(\z{mmsiaddrcfg}, \z{mmsiaddrcfgh}, \z{smsiaddrcfg}, and
\z{smsiaddrcfgh}), all of which retain their previous values.
(The root domain's \z{domaincfg} gets the value written when changing
field~CM, of course.)

The endianness of the memory-mapped registers of the original PLIC side
of the \mbox{Duo-PLIC} is not affected by the BE field of any
APLIC \z{domaincfg} register.
For most, maybe all versions of an original PLIC, the registers are
always little-endian.

The number of interrupt priority levels supported by the original PLIC
side of a \mbox{Duo-PLIC} is the same as that of the APLIC side
when in direct delivery mode.
Hence, the number of integer bits to store a priority number, IPRIOLEN,
is a constant for a \mbox{Duo-PLIC} (in the range 1 to 8 as usual),
independent of whether compatibility mode is enabled or disabled.

However, an original PLIC encodes priorities in the opposite order
from an IMSIC or APLIC, with number 1 being interpreted as the
lowest priority and larger numbers as higher priorities.
When a \mbox{Duo-PLIC} is acting as an original PLIC and reports to
a hart an interrupt priority number (see Chapter~\ref{ch:MSLevel}
concerning a hart's handling of interrupt priorities), the number
reported is negated to \emph{normalize} it to the convention of the
Advanced Interrupt Architecture.
If the original PLIC's priority number for an interrupt is~$p$,
the normalized priority number communicated to the target hart is
${\mbox{2}^\textrm{IPRIOLEN}-p}$.

### 2022-02-28 doc/src/IMSIC.tex
Adopt "APLIC" as common name for Advanced PLIC

\z{eidelivery} is a {\WARL} register that controls whether interrupts
from this interrupt file are delivered from the IMSIC to the attached
hart so they appear as a pending external interrupt in the hart's
\z{mip} or \z{hgeip} CSR.
Register \z{eidelivery} may optionally also support the direct delivery
of interrupts from a PLIC (Platform-Level Interrupt Controller) to the
attached hart.
Three possible values are currently defined for \z{eidelivery}:
\begin{displayLinesTable}[r@{\ }l]
0              & = Interrupt delivery is disabled \\
1              & = Interrupt delivery from the interrupt file is enabled \\
\z{0x40000000} & = Interrupt delivery from a PLIC is enabled (optional)\\
\end{displayLinesTable}

If \z{eidelivery} supports value \z{0x40000000}, then a specific PLIC
in the system may act as an alternate external interrupt controller for
the attached hart at the same privilege level as this interrupt file.
When \z{eidelivery} is \z{0x40000000}, the interrupt file functions the
same as though \z{eidelivery} is~0, and the PLIC replaces the interrupt
file in supplying pending external interrupts at this privilege level
at the hart.

@@

\begin{commentary}
\z{eidelivery} value \z{0x40000000} supports system software that is
oblivious to IMSICs and assumes instead that the external interrupt
controller is a PLIC.
Such software may exist either because it predates the existence of
IMSICs or because bypassing IMSICs is believed to reduce programming
effort.
\end{commentary}

>>>>

\z{eidelivery} is a {\WARL} register that controls whether interrupts
from this interrupt file are delivered from the IMSIC to the attached
hart so they appear as a pending external interrupt in the hart's
\z{mip} or \z{hgeip} CSR.
Register \z{eidelivery} may optionally also support the direct delivery
of interrupts from a PLIC (Platform-Level Interrupt Controller)
or APLIC (Advanced PLIC) to the attached hart.
Three possible values are currently defined for \z{eidelivery}:
\begin{displayLinesTable}[r@{\ }l]
0              & = Interrupt delivery is disabled \\
1              & = Interrupt delivery from the interrupt file is enabled \\
\z{0x40000000} & = Interrupt delivery from a PLIC or APLIC is enabled (optional)\\
\end{displayLinesTable}

If \z{eidelivery} supports value \z{0x40000000}, then a specific PLIC or APLIC
in the system may act as an alternate external interrupt controller for
the attached hart at the same privilege level as this interrupt file.
When \z{eidelivery} is \z{0x40000000}, the interrupt file functions the
same as though \z{eidelivery} is~0, and the PLIC or APLIC replaces the interrupt
file in supplying pending external interrupts at this privilege level
at the hart.

@@

\begin{commentary}
\z{eidelivery} value \z{0x40000000} supports system software that is
oblivious to IMSICs and assumes instead that the external interrupt
controller is a PLIC or APLIC.
Such software may exist either because it predates the existence of
IMSICs or because bypassing IMSICs is believed to reduce programming
effort.
\end{commentary}

### 2022-02-28 doc/src/MSLevel.tex
Adopt "APLIC" as common name for Advanced PLIC

The {\RISCV} Privileged Architecture defines several major identities
in the range 0--15 for interrupts at a hart, including machine-level
and supervisor-level external interrupts (numbers 11 and~9), machine-
and supervisor-level timer interrupts (7~and~5), and machine- and
supervisor-level software interrupts (3~and~1).
Beyond these major labels, the \emph{external} interrupts at each
privilege level are given secondary, minor identities by an external
interrupt controller such as a PLIC or IMSIC, distinguishing interrupts
from different devices or causes.
These minor identities for external interrupts were covered in Chapters
\ref{ch:IMSIC} and~\ref{ch:AdvPLIC} specifying the IMSIC and Advanced
PLIC components.

@@

Lastly, an optional facility lets software assign priorities to major
interrupts (such as the timer and software interrupts, and any local
interrupts) such that they may mix with the priorities set for external
interrupts by a PLIC or IMSIC.

@@

Even when a defined local interrupt is implemented, it is not
necessarily the case that all events matching the description for that
interrupt actually cause the local interrupt to be raised.
For example, a single {\RISCV} system might report bus errors in
multiple ways, depending on where in the system the error is detected.
Some bus errors might signal the standard local interrupt (number~30),
while others are reported as external interrupts, routed through a PLIC
or IMSIC.

@@

The width of priority numbers for external interrupts is
\emph{IPRIOLEN}.
This parameter is affected by the main external interrupt controller
for the hart, either a PLIC or an IMSIC.

For a PLIC, IPRIOLEN is in the range 1--8 as specified in
Chapter~\ref{ch:AdvPLIC} on the Advanced PLIC.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 & Interrupts with default & Machine external & Interrupts with default \\
 & priority above machine  & interrupts       & priority below machine \\
 & external interrupts     &                  & external interrupts \\
\hline
         & Priority number in & Priority number from & Priority number in \\
Priority & machine-level      & interrupt controller & machine-level \\
order    & \z{iprio} array    & (PLIC or IMSIC)      & \z{iprio} array \\
\hline
\hline
Highest  & 0                  &                      & \\
\cline{2-4}
         & 1                  & 1                    & 1 \\
         & 2                  & 2                    & 2 \\
         & $\cdots$           & $\cdots$             & $\cdots$ \\
         & 254                & 254                  & 254 \\
         & 255                & 255                  & 255 \\
\cline{2-4}
         &                    & 256 and above        & \\
         &                    & (IMSIC only)         & \\
\cline{2-4}
Lowest   &                    &                      & 0 \\
\hline
\end{tabular}
\end{center}
\caption{%
Effect of the machine-level \z{iprio} array on the priorities of
interrupts taken in \mbox{M-mode}.
For interrupts with the same priority number, the default order of
Section~\ref{sec:majorIntrs} prevails.%
}
\label{tab:intrPrios-M}
\end{table*}

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 & Interrupts with default  & Supervisor external & Interrupts with default \\
 & priority above supervisor & interrupts       & priority below supervisor \\
 & external interrupts       &                  & external interrupts \\
\hline
         & Priority number in & Priority number from & Priority number in \\
Priority & supervisor-level   & interrupt controller & supervisor-level \\
order    & \z{iprio} array    & (PLIC or IMSIC)      & \z{iprio} array \\
\hline
\hline
Highest  & 0                  &                      & \\
\cline{2-4}
         & 1                  & 1                    & 1 \\
         & 2                  & 2                    & 2 \\
         & $\cdots$           & $\cdots$             & $\cdots$ \\
         & 254                & 254                  & 254 \\
         & 255                & 255                  & 255 \\
\cline{2-4}
         &                    & 256 and above        & \\
         &                    & (IMSIC only)         & \\
\cline{2-4}
Lowest   &                    &                      & 0 \\
\hline
\end{tabular}
\end{center}
\caption{%
Effect of the supervisor-level \z{iprio} array on the priorities of
interrupts taken in \mbox{S-mode}.
For interrupts with the same priority number, the default order of
Section~\ref{sec:majorIntrs} prevails.%
}
\label{tab:intrPrios-S}
\end{table*}

>>>>

The {\RISCV} Privileged Architecture defines several major identities
in the range 0--15 for interrupts at a hart, including machine-level
and supervisor-level external interrupts (numbers 11 and~9), machine-
and supervisor-level timer interrupts (7~and~5), and machine- and
supervisor-level software interrupts (3~and~1).
Beyond these major labels, the \emph{external} interrupts at each
privilege level are given secondary, minor identities by an external
interrupt controller such as an APLIC or IMSIC, distinguishing interrupts
from different devices or causes.
These minor identities for external interrupts were covered in Chapters
\ref{ch:IMSIC} and~\ref{ch:AdvPLIC} specifying the IMSIC and
APLIC components.

@@

Lastly, an optional facility lets software assign priorities to major
interrupts (such as the timer and software interrupts, and any local
interrupts) such that they may mix with the priorities set for external
interrupts by a PLIC, APLIC, or IMSIC.

@@

Even when a defined local interrupt is implemented, it is not
necessarily the case that all events matching the description for that
interrupt actually cause the local interrupt to be raised.
For example, a single {\RISCV} system might report bus errors in
multiple ways, depending on where in the system the error is detected.
Some bus errors might signal the standard local interrupt (number~30),
while others are reported as external interrupts, routed through an APLIC
or IMSIC.

@@

The width of priority numbers for external interrupts is
\emph{IPRIOLEN}.
This parameter is affected by the main external interrupt controller
for the hart, whether a PLIC, APLIC, or IMSIC.

For an APLIC, IPRIOLEN is in the range 1--8 as specified in
Chapter~\ref{ch:AdvPLIC} on the APLIC.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 & Interrupts with default & Machine external & Interrupts with default \\
 & priority above machine  & interrupts       & priority below machine \\
 & external interrupts     &                  & external interrupts \\
\hline
         & Priority number in & Priority number from & Priority number in \\
Priority & machine-level      & interrupt controller & machine-level \\
order    & \z{iprio} array    & (APLIC or IMSIC)     & \z{iprio} array \\
\hline
\hline
Highest  & 0                  &                      & \\
\cline{2-4}
         & 1                  & 1                    & 1 \\
         & 2                  & 2                    & 2 \\
         & $\cdots$           & $\cdots$             & $\cdots$ \\
         & 254                & 254                  & 254 \\
         & 255                & 255                  & 255 \\
\cline{2-4}
         &                    & 256 and above        & \\
         &                    & (IMSIC only)         & \\
\cline{2-4}
Lowest   &                    &                      & 0 \\
\hline
\end{tabular}
\end{center}
\caption{%
Effect of the machine-level \z{iprio} array on the priorities of
interrupts taken in \mbox{M-mode}.
For interrupts with the same priority number, the default order of
Section~\ref{sec:majorIntrs} prevails.%
}
\label{tab:intrPrios-M}
\end{table*}

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
 & Interrupts with default  & Supervisor external & Interrupts with default \\
 & priority above supervisor & interrupts       & priority below supervisor \\
 & external interrupts       &                  & external interrupts \\
\hline
         & Priority number in & Priority number from & Priority number in \\
Priority & supervisor-level   & interrupt controller & supervisor-level \\
order    & \z{iprio} array    & (APLIC or IMSIC)     & \z{iprio} array \\
\hline
\hline
Highest  & 0                  &                      & \\
\cline{2-4}
         & 1                  & 1                    & 1 \\
         & 2                  & 2                    & 2 \\
         & $\cdots$           & $\cdots$             & $\cdots$ \\
         & 254                & 254                  & 254 \\
         & 255                & 255                  & 255 \\
\cline{2-4}
         &                    & 256 and above        & \\
         &                    & (IMSIC only)         & \\
\cline{2-4}
Lowest   &                    &                      & 0 \\
\hline
\end{tabular}
\end{center}
\caption{%
Effect of the supervisor-level \z{iprio} array on the priorities of
interrupts taken in \mbox{S-mode}.
For interrupts with the same priority number, the default order of
Section~\ref{sec:majorIntrs} prevails.%
}
\label{tab:intrPrios-S}
\end{table*}

### 2022-02-28 doc/src/VSLevel.tex
Adopt "APLIC" as common name for Advanced PLIC

In addition to an IMSIC at each hart, a virtual machine may also need
to see a PLIC.
However, unlike an IMSIC's ability to provide physical guest interrupt
files for virtual harts, a PLIC must be emulated for a virtual machine
by the hypervisor.

\begin{commentary}
The Advanced Interrupt Architecture does not currently include hardware
assistance for virtualizing a PLIC.
For small numbers of harts, such hardware would be substantially larger
than that required to implement guest interrupt files for an IMSIC.
It is assumed that most high-performance I/O can be done through
devices that can send MSIs directly to guest interrupt files (such as
devices attached through a PCI Express interconnect).
For the types of devices whose interrupts must go through a (virtual)
PLIC, the overhead cost of emulating the PLIC is expected to be less
significant.
\end{commentary}

@@

In the special case that an emulated Advanced PLIC for a virtual
machine has a wired interrupt source that equates to an actual
interrupt source of a real Advanced PLIC, if software running in this
virtual machine configures its virtual PLIC to forward interrupts from
that source as MSIs to a specific virtual hart, the hypervisor can
configure the real PLIC to forward the actual interrupts directly as
MSIs to the virtual hart's guest interrupt file.
In this way, although the hypervisor must trap and emulate the virtual
machine's memory accesses that configure the forwarding of interrupts
at the virtual PLIC, the interrupts themselves can be converted
automatically into real MSIs for the guest interrupt file, without the
hypervisor being invoked for each arriving interrupt.

@@

To ensure proper support for interrupts, two conditions must be met
before a hypervisor may allow a guest OS running in a virtual machine
to directly control a physical device that sends MSIs:
First, each virtual hart must have a guest interrupt file assigned to
it, giving each its own apparent IMSIC within the virtual machine.
Second, interrupts from the device must be signaled by wire through
an Advanced PLIC that can translate these interrupts into MSIs, or the
system must have an I/O MMU that can translate the addresses of MSI
memory writes made by the device itself.

@@

\item
Modify the relevant translation tables at all I/O MMUs so that MSIs for
this virtual interrupt file are now sent to the new physical interrupt
file.
Likewise, if any interrupts at a PLIC are forwarded by MSIs to the old
interrupt file, reconfigure the PLIC to send them to the new interrupt
file.
As needed, synchronize with all I/O MMUs and PLICs to ensure that no
straggler MSIs will arrive at the old interrupt file after this step.
Synchronizing with an Advanced PLIC can be accomplished using the
algorithm of Section~\ref{sec:AdvPLIC-MSISync}.

>>>>

In addition to an IMSIC at each hart, a virtual machine may also need
to see a PLIC or APLIC.
However, unlike an IMSIC's ability to provide physical guest interrupt
files for virtual harts, a PLIC or APLIC must be emulated for a virtual machine
by the hypervisor.

\begin{commentary}
The Advanced Interrupt Architecture does not currently include hardware
assistance for virtualizing an APLIC.
For small numbers of harts, such hardware would be substantially larger
than that required to implement guest interrupt files for an IMSIC.
It is assumed that most high-performance I/O can be done through
devices that can send MSIs directly to guest interrupt files (such as
devices attached through a PCI Express interconnect).
For the types of devices whose interrupts must go through a (virtual)
APLIC, the overhead cost of emulating the APLIC is expected to be less
significant.
\end{commentary}

@@

In the special case that an emulated APLIC for a virtual
machine has a wired interrupt source that equates to an actual
interrupt source of a real APLIC, if software running in this
virtual machine configures its virtual APLIC to forward interrupts from
that source as MSIs to a specific virtual hart, the hypervisor can
configure the real APLIC to forward the actual interrupts directly as
MSIs to the virtual hart's guest interrupt file.
In this way, although the hypervisor must trap and emulate the virtual
machine's memory accesses that configure the forwarding of interrupts
at the virtual APLIC, the interrupts themselves can be converted
automatically into real MSIs for the guest interrupt file, without the
hypervisor being invoked for each arriving interrupt.

@@

To ensure proper support for interrupts, two conditions must be met
before a hypervisor may allow a guest OS running in a virtual machine
to directly control a physical device that sends MSIs:
First, each virtual hart must have a guest interrupt file assigned to
it, giving each its own apparent IMSIC within the virtual machine.
Second, interrupts from the device must be signaled by wire through
an APLIC that can translate these interrupts into MSIs, or the
system must have an I/O MMU that can translate the addresses of MSI
memory writes made by the device itself.

@@

\item
Modify the relevant translation tables at all I/O MMUs so that MSIs for
this virtual interrupt file are now sent to the new physical interrupt
file.
Likewise, if any interrupts at an APLIC are forwarded by MSIs to the old
interrupt file, reconfigure the APLIC to send them to the new interrupt
file.
As needed, synchronize with all I/O MMUs and APLICs to ensure that no
straggler MSIs will arrive at the old interrupt file after this step.
Synchronizing with an APLIC can be accomplished using the
algorithm of Section~\ref{sec:AdvPLIC-MSISync}.

### 2022-02-28 doc/src/intro.tex
Adopt "APLIC" as common name for Advanced PLIC

This document specifies the Advanced Interrupt Architecture for
{\RISCV}, consisting of:
(a)~an extension to the standard Privileged Architecture for {\RISCV}
harts specified in Volume~II of The {\RISCV} Instruction Set Manual;
(b)~two standard interrupt controllers for {\RISCV} systems, an
Advanced Platform-Level Interrupt Controller (Advanced PLIC) and an
Incoming Message-Signaled Interrupt Controller (IMSIC); and
(c)~requirements on other system components concerning interrupts.

@@

\item
For wired interrupts, define a new Platform-Level Interrupt Controller
(Advanced PLIC) that has an independent control interface for each
level of privilege (such as {\RISCV} machine and supervisor levels),
and that can convert wired interrupts into MSIs for systems supporting
MSIs.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|c|l|}
\hline
   & Maximum      & \multicolumn{1}{c|}{Requirements} \\
\hline
\hline
Physical harts
   & 16,384       & \\
\hline
Active virtual harts having direct control of \quad
   & 31 for RV32, & {\RISCV} hypervisor extension; \\
\ a device, per physical hart
   & 63 for RV64  & \ IMSICs with guest interrupt\\
   &              & \ files; and an I/O MMU \\
\hline
Idle (swapped-out) virtual harts having \quad
   & potentially  & An I/O MMU with support \\
\ direct control of a device, per physical hart \quad
   & thousands    & \ for memory-resident \\
   &              & \ interrupt files \\
\hline
Wired interrupts at a single Advanced PLIC \quad
   & 1023         & \\
\hline
Distinct identities usable for MSIs at each \quad
   & 2047         & IMSICs \\
\ hart (physical or virtual)
   &              & \\
\hline
\end{tabular}
\end{center}
\caption{%
Absolute limits on the numbers of harts and interrupt identities in a
system.
Individual implementations are likely to have smaller limits.%
}
\label{tab:overallLimits}
\end{table*}

@@

When {\RISCV} harts do not have Incoming MSI Controllers, external
interrupts are signaled to harts through dedicated wires.
In that case, an \emph{Advanced Platform-Level Interrupt Controller}
(Advanced PLIC) acts as a traditional central hub for interrupts,
routing and prioritizing external interrupts for each hart as
illustrated in Figure~\ref{fig:intrsWithoutIMSICs}.
Interrupts may be selectively routed either to machine level or to
supervisor level at each hart.
The Advanced PLIC is specified in Chapter~\ref{ch:AdvPLIC}.

@@

\begin{commentary}
If harts implement the hypervisor extension, it is a topic of ongoing
study whether an Advanced PLIC should be allowed to route external
interrupts to be the \emph{guest external interrupts} of the hypervisor
extension, permitting the delivery of interrupts directly to virtual
machines without the need for each signaled interrupt to be handled at
the hypervisor level.
For now, we assume that systems that need direct signaling of external
interrupts to virtual machines will have IMSICs.
\end{commentary}

@@

When the harts in a {\RISCV} system have IMSICs, the system will
normally still contain an Advanced PLIC, but its role is changed.
Instead of signaling interrupts to harts directly by wires as in
Figure~\ref{fig:intrsWithoutIMSICs}, an Advanced PLIC converts incoming
wired interrupts into MSI writes that are sent to harts via their IMSIC
units.
Each MSI is sent to a single target hart according to the PLIC's
configuration set by software.

@@

The Advanced Interrupt Architecture adds considerable support
for \emph{local interrupts} at a hart, whereby a hart essentially
interrupts itself in response to asynchronous events, usually errors.
Local interrupts remain contained within a hart (or close to it),
so like standard {\RISCV} timer and software interrupts, they do not
pass through an Advanced PLIC or IMSIC.

@@

Interrupts from most I/O devices are conveyed to a hart by the
\emph{external interrupt controller} for the hart, which is either the
hart's IMSIC (Figure~\ref{fig:intrsWithIMSICs}) or an Advanced PLIC
(Figure~\ref{fig:intrsWithoutIMSICs}).
As Table~\ref{tab:interruptIdents} shows, external interrupts
at a given privilege level all share a single major identity
number:  11~for machine level, 9~for supervisor level, and 10 for
\mbox{VS-level}.
External interrupts from different causes are distinguished from one
another at a hart by their \emph{minor identity numbers} supplied by
the external interrupt controller.

>>>>

This document specifies the Advanced Interrupt Architecture for
{\RISCV}, consisting of:
(a)~an extension to the standard Privileged Architecture for {\RISCV}
harts specified in Volume~II of The {\RISCV} Instruction Set Manual;
(b)~two standard interrupt controllers for {\RISCV} systems, an
Advanced Platform-Level Interrupt Controller (APLIC) and an
Incoming Message-Signaled Interrupt Controller (IMSIC); and
(c)~requirements on other system components concerning interrupts.

@@

\item
For wired interrupts, define a new Platform-Level Interrupt Controller
(the Advanced PLIC, or APLIC)
that has an independent control interface for each
level of privilege (such as {\RISCV} machine and supervisor levels),
and that can convert wired interrupts into MSIs for systems supporting
MSIs.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|c|l|}
\hline
   & Maximum      & \multicolumn{1}{c|}{Requirements} \\
\hline
\hline
Physical harts
   & 16,384       & \\
\hline
Active virtual harts having direct control of \quad
   & 31 for RV32, & {\RISCV} hypervisor extension; \\
\ a device, per physical hart
   & 63 for RV64  & \ IMSICs with guest interrupt\\
   &              & \ files; and an I/O MMU \\
\hline
Idle (swapped-out) virtual harts having \quad
   & potentially  & An I/O MMU with support \\
\ direct control of a device, per physical hart \quad
   & thousands    & \ for memory-resident \\
   &              & \ interrupt files \\
\hline
Wired interrupts at a single APLIC \quad
   & 1023         & \\
\hline
Distinct identities usable for MSIs at each \quad
   & 2047         & IMSICs \\
\ hart (physical or virtual)
   &              & \\
\hline
\end{tabular}
\end{center}
\caption{%
Absolute limits on the numbers of harts and interrupt identities in a
system.
Individual implementations are likely to have smaller limits.%
}
\label{tab:overallLimits}
\end{table*}

@@

When {\RISCV} harts do not have Incoming MSI Controllers, external
interrupts are signaled to harts through dedicated wires.
In that case, an \emph{Advanced Platform-Level Interrupt Controller}
(APLIC) acts as a traditional central hub for interrupts,
routing and prioritizing external interrupts for each hart as
illustrated in Figure~\ref{fig:intrsWithoutIMSICs}.
Interrupts may be selectively routed either to machine level or to
supervisor level at each hart.
The APLIC is specified in Chapter~\ref{ch:AdvPLIC}.

@@

\begin{commentary}
If harts implement the hypervisor extension, it is a topic of ongoing
study whether an APLIC should be allowed to route external
interrupts to be the \emph{guest external interrupts} of the hypervisor
extension, permitting the delivery of interrupts directly to virtual
machines without the need for each signaled interrupt to be handled at
the hypervisor level.
For now, we assume that systems that need direct signaling of external
interrupts to virtual machines will have IMSICs.
\end{commentary}

@@

When the harts in a {\RISCV} system have IMSICs, the system will
normally still contain an APLIC, but its role is changed.
Instead of signaling interrupts to harts directly by wires as in
Figure~\ref{fig:intrsWithoutIMSICs}, an APLIC converts incoming
wired interrupts into MSI writes that are sent to harts via their IMSIC
units.
Each MSI is sent to a single target hart according to the APLIC's
configuration set by software.

@@

The Advanced Interrupt Architecture adds considerable support
for \emph{local interrupts} at a hart, whereby a hart essentially
interrupts itself in response to asynchronous events, usually errors.
Local interrupts remain contained within a hart (or close to it),
so like standard {\RISCV} timer and software interrupts, they do not
pass through an APLIC or IMSIC.

@@

Interrupts from most I/O devices are conveyed to a hart by the
\emph{external interrupt controller} for the hart, which is either the
hart's IMSIC (Figure~\ref{fig:intrsWithIMSICs}) or an APLIC
(Figure~\ref{fig:intrsWithoutIMSICs}).
As Table~\ref{tab:interruptIdents} shows, external interrupts
at a given privilege level all share a single major identity
number:  11~for machine level, 9~for supervisor level, and 10 for
\mbox{VS-level}.
External interrupts from different causes are distinguished from one
another at a hart by their \emph{minor identity numbers} supplied by
the external interrupt controller.
